------------------------------------------------------------------------

------------------------------------------------------------------------

```{r}
#| label: Read in Data

library(here)
library(readxl)

humorincog <- read_excel(here("data/pone.0211618.s002.xlsx"))

# replicate correlation in Table 3 between Negative affectivity (naff) and General Positive Aspects (pos3)
# Reported values: -0.26 (0.011)
cor1 <-cor.test(humorincog$pos3, humorincog$naff)
cor1$estimate 
cor1$p.value

# replicate linear regression coefficient Table 3
# The six eligible reappraisal sub-strategies (three positive  reinterpretation strategies (pos3, pos7, pos8), 
# three de-emphasising strategies (rel10, rel11, rel14)) 
# were used as predictors of Antagonism (anta)
# Estimated Coefficient of Alternative Explanation (rel14)
# Reported Values: 0.12 (0.15)

anta_reg <- lm(data=humorincog, anta~pos3+pos7+pos8+rel10+rel11+rel14)
summary(anta_reg)
```

## What is R

## Why use R?

-   Stuff from <https://www.discovr.rocks/lover/>
-   Usage stats (Rexer analytics' surveys etc)
-   Reproducibility
-   Cutting edge

## Using R

## Example analysis

To demonstrate R's extensive analytic capability we use data from the
study conducted by Perchtold et. al. (2019), “Humorous cognitive
reappraisal: More benign humour and less "dark" humour is affiliated
with more adaptive cognitive reappraisal strategies.” In this study 95
qualified participants aged 17 to 35 years completed the Reappraisal
Inventiveness Test (RIT), humorous adaptation, and questionnaires.
Participants were confronted with 4 anxiety-provoking situations and
were asked to write as many humorous ways as they could think of to
diminish anxiety. Each of the humorous explanations (or reappraisals)
was classified into RIT categories; which were further classified into
either positive reinterpretation strategy (situation-focused
reappraisal) or a de-emphasizing strategy (self-focused reappraisal).
Participants also completed the Center for Epidemiology Studies
depression scale (CES-D, variable name ads), resulting in an overall
score from 1 to 48, with higher scores indicating a greater number of
self-reported depressive experiences.

### One-Sample T-Test

A first analysis determines whether the participants do not show
depressive symptoms on average which is reflected as a CES-D score of 15
or lower (Stein 2014). The threshold of 15 was chosen because a score of
16 or is interpreted as displaying depressive symptoms. This hypothesis
can be tested using a one-sample t-test.

H0: true mean CES-D is equal to 15

H1: true mean CES-D is less than 15

In R we perform a t-test of this hypothesis using the t.test() function
which is available in base R. Mu is the constant which is compared to
the sample mean and in this case mu is equal to 15. The alternative
hypothesis is set to "less" to perform a one-tailed test. The object
produced from the t-test (called depression) is printed.

```{r}
#| label: One-sample t-test

depression <- t.test(x=humorincog$ads, mu=15, alternative="less" )
depression
```

Upon examination of the output the p-value for this test is 0.7502 which
is greater than 𝛼=0.05. We have insufficient evidence to reject the null
hypothesis and the decision is to Fail to Reject the null hypothesis. We
can conclude at the 5% level of significance that the true mean CES-D
value is not less than 15.

### Correlation Coefficient

In two-way comparisons Perchtold and colleagues correlate self-reported
depressive experiences to each of the three sub-strategies within the
positive reinterpretation strategy (general positive aspects (pos3),
worst-case comparison (pos7), disadvantage as advantage (pos8)) and each
of the three sub-strategies within the de-emphasizing category
(alternative explanation (rel10), trivializing the problem (rel11),
handing over responsibility (rel14)). (Perchtold Table 2). The columns
of interest for the analysis are extracted and then a simple correlation
matrix can be created using the cor() function. The round() function is
wrapped around the correlations to display 3 digits after the decimal
and streamline the display. The default method is Pearson, and the
Kendall and Spearman correlations can be produced using method="kendall"
or "spearman".

```{r}
#| label: Correlation

humorincog_c<-humorincog[c("ads","pos3","pos7","pos8","rel10","rel11","rel14")]
round(cor(humorincog_c),3)

```

Standard errors (or p-values) for the correlations can be produced for
each pair using the corr.test() function (as in Perchtold Table 2).

```{r}
#| label: Correlation with P value

library(psych)

ads_corr <- corr.test(humorincog_c,adjust="none")
ads_corr

```

The object ads_corr contains the correlation matrix, followed by the
sample size and the unadjusted p-values of the correlations.

```         
{r}
```

### Multiple Linear Regression

Following Perchtold et al. (2019) and using multivariate linear
regression, we can test whether three sub-strategies within the positive
reinterpretation strategy (general positive aspects (pos3), worst-case
comparison (pos7), disadvantage as advantage (pos8)) and three
sub-strategies within the de-emphasizing category (alternative
explanation (rel10), trivializing the problem (rel11), handing over
responsibility (rel14)) are predictive of self-reported depressive
experiences measured via the CES-D depression scale. \[These results are
reported in Table 2 of Perchtold.\]

Several of the research questions that can be answered using multiple
linear regression:

1.  Do *any* of the six reappraisal strategies have a linear
    relationship to self-reported depressive experience? (overall test
    of the model)

2.  Which of the six strategies explain variation in self-reported
    depressive experience? (test the estimated parameter for each
    strategy = 0)

3.  What is the size of the relationship between a specific strategy and
    self-reported depressive experience? (confidence interval for a
    parameter)

4.  What is the predicted CES-D depression score for a person with a
    given set of the six strategies? (prediction interval)

In R we answer these questions by estimating a linear regression using
the lm( ) function. We designate the data frame which is stored in
object humorincog and specify the model using a formula with the
dependent variable on the left side of a tilde (\~) and the explanatory
variables on the right side of the tilde, separated by plus (+) signs.

ads \~ pos3+pos7+pos8+rel10+rel11+rel14

The estimated model can be written to an object, called ads_reg in this
case, and the model results are processed using the summary( ) function.

```{r}
#| label: Multiple Regression

ads_reg <- lm(data=humorincog, ads~pos3+pos7+pos8+rel10+rel11+rel14)
summary(ads_reg)


```

The regression output provides answers to our research questions. \[Too
much detail? If not, each would be answered using the values in the
output\]

1.  Do *any* of the six reappraisal strategies have a linear
    relationship to self-reported depressive experience? (overall test
    of the model)

2.  Which of the six strategies explain variation in self-reported
    depressive experience? (test the estimated parameter for each
    strategy = 0)

3.  What is the size of the relationship between a specific strategy and
    self-reported depressive experience? (confidence interval for a
    parameter)

4.  What is the predicted CES-D depression score for a person with a
    given set of the six strategies? (prediction interval)

Producing Tables of Output

Often we would like to produce well-formatted tables of regression
output for reporting or publication, and packages such as broom() make
this process much more efficient. The tidy() function standardizes the
output from the model and the kable() function generates a table of the
contents.

```{r}
#| label: Regression tables 

library(knitr)
library(broom)
tidy(ads_reg) |> kable(digits = 2)
```

Classes and Lists

All of the statistical tests above produce useful data elements that are
unseen in the output that is typically produced by printing the object,
by using summary(), or by running the object through tidy() and broom().
For example, the output produced in an object created by the lm()
function includes useful data such as all of the residuals and fitted
values. These do not need to be calculated; you just need to learn how
to extract them.

The first step is to write your output to an object. The object will
have standardized names for variables and a standardized format. It will
be stored in what is called a *list* object which is an object that
contains objects of different types. Where a data frame or tibble has to
have a fixed number of rows across all columns, a list is not limited to
that structure. Within the list an object can be a character vector of
length 1 or a numeric array with dimensions nxk.

All lists produced by lm() include the following objects:\[usful?\]

In R, each object created by lm() in the structure listed above is
designated as a class "lm" object. Classes allow creation of a template
to store information, and just as importantly, to extract information.
Every "lm" class object will contain residuals, fitted values, etc...
and users can write functions or packages to manipulate data produced as
a class-\[insert class name\] object such as a class-lm object. Packages
such as broom are equipped to check the class of an object and "tidy" it
to an even more standardized format. But we don't need to tell broom
what class an object is, the object itself provides that information.

\[How classes are useful to the R user\]

## References

Perchtold CM, Weiss EM, Rominger C, Feyaerts K, Ruch W, Fink A, et al.
(2019) Humorous cognitive reappraisal: More benign humour and less
"dark" humour is affiliated with more adaptive cognitive reappraisal
strategies. PLoS ONE 14(1): e0211618.
<https://doi.org/10.1371/journal.pone.0211618>

Stein J, Luppa M, Mahnke J, Weyerer S, Schomerus G, Riedel-Heller SG.
Depressionsscreening am Telefon mittels der Allgemeinen Depressionsskala
(ADS) \[Screening for depression by telephone using the German version
of the Center for Epidemiological Studies Depression Scale (CES-D)\].
Psychiatr Prax. 2014 Apr;41(3):135-41. German. doi:
10.1055/s-0033-1343176. Epub 2013 May 13. PMID: 23670784.
